---
title: "Replicating Scotland's NCAI"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{replicating_scotlands_ncai}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(openNCAI)
```

## Introduction

Scotland's **Natural Capital Asset Index (NCAI)** is designed by NatureScot to monitor the value of Scotland's Natural Capital Assets. This vignette, written to demonstrate the package to our NatureScot colleagues, demonstrates how to use the `openNCAI` package to replicate the index calculations.

## Data Preparation

To begin, we import data from NatureScot's original spreadsheet containing the index calculations to year 2022. A copy of the spreadsheet is included in the `extdata` directory. The get_ns_data() command is not exported with the publicly available package, but can be accessed with the 'triple colon' notation, like this: `openNCAI:::get_ns_data()`.

NOTE TO SELF - we also have the option of exporting but hiding from the help by adding `@keywords internal` to the Roxygen comments. 

```{r import_data}
# Locate the template within the package
path <- system.file("data-raw", "ncai_corrected.xlsx", package = "openNCAI")

# Import all trees, scores, and matrices
# The year_list represents the temporal coverage of the account
# ns_data <- openNCAI:::import_ns_data(ns_sheets_path, year_list = 2000:2022)
```

The function imports a list of objects:

```{r show_nsdata_list}
# print(names(ns_data))
```
