---
title: "replicating_scotlands_ncai"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{replicating_scotlands_ncai}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

The 'openNCAI' package provides a reproducible pipeline to calculate a regional natural capital assets index (NCAI) using the method designed by NatureScot to calculate Scotland's NCAI.

To illustrate this we use the built-in assets provided by NatureScot.

```{r setup}
library(openNCAI)
```

## Step 1: Accessing Package Data

The following NatureScot data objects are bundled with the package:

```{r data-table, echo = FALSE}
# 1. Capture the data list
d <- as.data.frame(data(package = "openNCAI")$results[, c("Item", "Title")])

# 2. Rename columns for a cleaner look
colnames(d) <- c("Object Name", "Description")

# 3. Render as a formatted table that wraps text
knitr::kable(d, caption = "Available NatureScot Datasets")
```

Because these are built in to the package we can access the data objects directly. For example, we can show a snippet of the habitat extent data by calling the object 'ns_habitat_extent':

```{r inspect_measurement_data}
head(ns_habitat_extent[,1:3])
```

## Step 2: Understanding The Required Data Inputs

openNCAI requires three types of data:

1. **Metadata** which define the natural habitats under consideration, the ecosystem services they provide, and the years across which the index is to be calculated,

2. **Environmental measures** recording the extent and condition of the natural habitats over the years,

3. **Weighting scores** - decided by experts at the beginning of the process; these denote the relative importance of different ecosystem services, the potential of different habitats to provide each each, and the salience of the available habitat condition indicators to represent flow of services from habitats.

Let's


